Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-09-11 paper/paper.tex
small revision on abstract and intro

\begin{abstract}
We introduce a new approach to compute various properties of independent sets by encoding the problems into tensor networks with different tensor element algebra.
The type of graph problems that can be computed using this method includes: independence number, the number of independent sets, independence polynomial, maximal independence polynomial, and enumeration of the maximum independence set configurations.
Although the computational complexity inevitably scales exponentially with the graph size, the performance of our method using tensor network contraction with generic programming is on par or outperforms state-of-the-art, sophisticated methods;
on the other hand, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
\end{abstract}

@@

Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for the MIS requires time $O(2^{|V|})$. More efficient exact algorithms include the branching algorithms and dynamic programming.
Sophisticated branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^n n ^{O(1)}$~\cite{Xiao2017}.
Dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.

The independent set problem is closely related to the clique problem and the vertex cover problem~\cite{Moore2011};
more concretely, an MIS of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
Together, these problems find a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}.
Besides the standard MIS problem, there are many other interesting problems pertaining to independent sets, such as the number of independent sets,
independence polynomial, maximal independence polynomial, and enumeration of optimal and sub-optimal configurations.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions and to identify harder graphs in an ensemble of graphs \red{cite experiment}.

In this work, we introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map different problems into generic tensor network contraction with various tensor element algebra.
Our approach does not necessarily provide a better time complexity compared to dynamic programming,
but the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts.
We benchmark our algorithms, the implementation of which benefits from recent advances in tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021}.
Lastly, we provide a few examples and show that the toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.

>>>>

\begin{abstract}
We introduce a new approach to compute various properties of independent sets by encoding the problems and the graph into tensor networks with different tensor element algebra.
The type of graph problems that can be computed using this method includes: independence number, the number of independent sets, independence polynomial, maximal independence polynomial, and both counting and enumeration of all maximum independence sets, the large-size independent sets, and all maximal independent sets.
Although the computational complexity inevitably scales exponentially with the graph size, our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
\end{abstract}

@@

Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for the MIS requires time $O\left(2^{|V|} \right)$. More efficient exact algorithms include the branching algorithms and dynamic programming.
Sophisticated branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}.
Dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.

The independent set problem is closely related to the clique problem and the vertex cover problem~\cite{Moore2011};
more concretely, an MIS of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
Together, these problems find a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}.
Besides the standard problem of finding the independence number and an MIS, there are many other interesting problems pertaining to independent sets, such as the number of independent sets,
independence polynomial, maximal independence polynomial, and enumeration of MIS or maximal independent set configurations.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions and to identify harder graphs in an ensemble of graphs \red{cite experiment}.

In this work, we introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map different problems into generic tensor network contraction with various tensor element algebra.
Our approach does not necessarily provide a better time complexity compared to dynamic programming,
but the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on CPUs and GPUs, the implementation of which benefits from recent advances in tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021}.
Lastly, we provide a few examples and show that the toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
